The tree line ecotone punctuates the abundance and turnover of plants, pollinators, and their interactions

### 2021-04-14 ms_1/ms_1.Rmd
set up null model approach to preference analysis

---
title: "The tree line ecotone punctuates the elevational turnover of plants, pollinators, and their interactions"

@@

Data are not yet provided but will be permanently archived on Dryad if the paper is accepted for publication. This submission uses novel code, which is provided as Supporting Information to be evaluated as part of the submission. If the paper is accepted, the code will be archived on Dryad alongside the data.

@@

Mountains are unique ecological theaters in which extrinsic climate dynamics interact with intrinsic climate gradation to shape the distributions of species and biotic interactions [@Telwala2013-sc; @CaraDonna2014-nq; @Miller-Struttmann2014-wz; @Rafferty2020-aj]. The rate of recent climate change, however, is forcing species to adapt rapidly within the constraints of gene pools shaped by cooler climates [@Visser2008-ct]. As temperatures rise, species are expected to track suitable conditions to higher elevations, and such effects are already measurable across a broad range of taxa [@Lenoir2015-tj]. Elevational range shifts, however, are not merely abstract translations along the linear lapse of temperature generated by mountain slopes; they are movements *in situ*, involving the advent, extinction, and reorganization of biological interactions [@Tylianakis2008-fi; @Blois2013-dz] as well as the concomitant variation of abiotic conditions other than temperature [@Korner1995-mh; @Hodkinson2005-af]. Thus, when species move along continuous, linear climate gradients, they can experience punctuated and nonlinear ecological transitions. A salient example is the tree line ecotone that marks the elevational transition between forest and grassland. At this relatively discrete band within a linear elevational temperature gradient, plant communities exhibit rapid species turnover [@Descombes2017-id] and declining diversity [@Becker2007-cg] as abiotic conditions become more extreme away from the shelter of tree canopy [@Slatyer1992-ir].

@@

Although per-sample effort was standardized across all years, the total number of samples (i.e. site-dates) varied: 153 in 2010, 212 in 2011, and 539 in 2012. Since the start of field sampling each year was determined by snowmelt, lower sites were sampled as early as April while higher sites could not be sampled until June or July. This resulted in a temporally staggered pattern of sampling in 2010 and 2012, while in 2011 a combination of early snowmelt and a late start to field activities resulted in sampling that was approximately aligned in time across the whole elevation gradient. See Supplemental Material for details of sampling patterns.

@@

Bumble bee abundance was quantified as the total number of recorded floral visits per bumble bee species per site-date. To weight floral abundance based on the foraging preferences of each bumble bee species in our network, we scored each plant genus in our data set as either visited or not visited by each bumble bee species. We then calculated the total floral resource availability for each bumble bee species for each site-date by summing the observed flower cover of *visited* genera. We opted to work at the genus level based on the reasoning that if a bumble bee species visits one member of a given genus, other members of the same genus should also be considered potential floral hosts. This approach was intended to dampen the effects of false non-detection, particularly for rare bumble bee species whose diet breadth would tend to be underestimated simply due to sparsity of observations [@Williams2005-yp]. We also opted to treat the cryptic *B. terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be visually distinguished. We similarly pooled representatives of the parasitic *Psithyrus* subgenus --- *B. barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into the species group *B. psithryus*.

@@

We analyzed the abundance of bumble bees and floral resources, respectively, using hierarchical generalized additive models (HGAMs) [@Wood2011-pg; @Pedersen2019-kf]. To focus on elevational patterns, we chose to collapse the temporal variation of abundance within sites into summary values. For bumble bees, we modeled *peak* abundance at each site as a smooth response to `elevation`, with bumble bee `species` as a smoothing factor, `management` (grazing, mowing, or unmanaged) as a fixed effect, and `species:year` as a random intercept term (exploratory analysis indicated that species varied differently across years). For floral resources, we modeled *mean* flower cover, weighted per bumble bee species (i.e. the sum of *visited* genera), as a smooth response to `elevation`, with `management` as a fixed effect, and `year` and bumble bee `species` as random intercept terms. For the bumble bee model, we used the negative binomial distribution family (log link function), and for the floral model we used a Tweedie family distribution (log link function). Both models follow the "type I" model form described in @Pedersen2019-kf.

@@

For our analyses of species and interaction β-diversity, we pooled all samples (including across years) within sites, since the sparse observations of individual sampling events tended to inflate β-diversity and its variance to a degree that impaired model fitting and interpretation. After pooling, we calculated the species and interaction β-diversity between all pairs of sites using the binary Jaccard index and partitioned total interaction β-diversity into its components of species turnover (changes in species composition) and interaction rewiring (interaction turnover despite species constancy) [@Novotny2009-my; @Poisot2012-fk]. As an initial analysis, we visualized the relationship between site-wise elevation difference and each metric of β-diversity, fitting each response with a binomial regression smooth.

@@

To investigate the relationship between elevation and β-diversity more deeply, we performed a second analysis using generalized dissimilarity modeling (GDM) [@Ferrier2007-ji] to analyze the elevational variation in the β-diversity of bumble bees, flora, and (unpartitioned) interactions, respectively. In addition to estimating the relationship between differences in a response variable and differences in a predictor variables, GDM captures the slope of this relationship over the range of the predictor variable, revealing potential variation in the amount of change in the response induced by a given change in the predictor [@Ferrier2007-ji]. This enabled us to ask whether the relationship between β-diversity and elevation exhibits thresholds or other nonlinearities. We also included site management in each model by encoding management classes as a series of binary dummy variables, following @Ferrier2007-ji. To control for potential confounding effects of the number of sampling days per site, sampling days was included as a covariate. The importance of the variables in each model were assessed by a permutation test in which variable importance is defined as the percent loss of explained deviance when a given variable is permuted in isolation [@Fitzpatrick-2021aa].

@@

All analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. GAM analyses were performed with packages `mgcv` [@Wood_2017] and `mgcViz` [@Fasiolo2018aa]. β-diversity was calculated with the package `vegan` [@Oksanen_2019] and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Calculation and partitioning of interaction β-diversity were performed with the package `bipartite` [@Dormann2008-aa; Dormann2009-aa]. GDM analysis of β-diversity was performed with the package `gdm` [@Fitzpatrick-2021aa]. Annotated `R` code is available in the Supplementary Material.

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species, 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. *B. pascuorum* ("pasc"), *B. pratorum* ("prat"), *B. soroeensis* ("soro"), *B. terrestris/lucorum* ("telu"), and *B. wurflenii* ("wurf") accounted for the bulk of overall bumble bee abundance, while *B. hortorum* ("hort"), *B. jonellus* ("jone"), *B. psithyrus* ("psit"), *B. monticola* ("mont"), *B. mendax* ("mend"), *B. mucidus* ("muci"), *B. pyrenaeus* ("pyre"), and *B. gerstaeckeri* ("gers") were present at moderate to low abundance. *B. hypnorum*, *B. humilis*, and *B. lapidarius* were recorded only sporadically, so we omitted them from abundance analyses but retained them for β-diversity analyses. See Supplementary Material for a summary of species-wise abundances.

Floral surveying yielded a total of 352 plant species (192 genera and 52 families). Of these, 163 species (111 genera and 37 families) were observed to be visited by bumble bees. Eight species --- *Rubus idaeus*, *Rosa canina*, *Juniperus communis*, *Larix decidua*, *Salix* sp., *Caltha palustris*, *Pulmonaria officinalis*, and *Rheum barbarum* --- were recorded in visitation data but not recorded during floral surveying. Each accounted for no more than 4 visits in total over the three years of our study, and they were omitted from the analysis of floral abundance.

@@

HGAM analysis revealed that the response of peak abundance to elevation differed significantly (p ≤ 0.02) from a flat line for all bumble bee species **(Figure 2A)**. *B. mendax*, *B. monticola*, *B. mucidus*, *B. pyrenaeus*, and *B. soroeensis* exhibited a positive response to elevation and peaked in abundance above the tree line, consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. In contrast, the abundance of *B. hortorum*, *B. pascorum*, *B. psithyrus*, and *B. terrestris/lucorum* declined with elevation. This effect was especially pronounced in *B. pascuorum*, likely because its exceptionally long nesting cycle is incompatible with the short growing season at high elevations [@Neumayer2009-gx]. *B. gerstaeckeri*, *B. jonellus*, and *B. wurflenii* peaked at mid-elevation below the tree line, while *B. pratorum* was unique in exhibiting a sharp peak of abundance *at* the tree line. Our model explained 71% of total variance (adjusted R^2^ = 0.45). Neither mowing (p = 0.08) nor non-management (p = 0.95) had a significant effect on bumble bee abundance relative to the reference level of grasing management. The random effect of species:year was also nonsignificant (p = 0.52). See Supplemental Material for tabulated model output and diagnostics.

Mean floral resource abundance likewise responded significantly (p << 0.0001) to elevation for all bumble bee species **(Figure 2B)**. *B. jonellus*, *B. mendax*, *B. monticola*, *B. mucidus*, *B. soroeensis*, and *B. wurflenii* exhibited bimodal patterns of flora resource abundance, with peaks around 1000 and 1600 m.s.l, the latter peak corresponding closely to the location of the tree line. *B. pyrenaeus* exhibited a unimodal floral resource pattern peaking near 1600 m.s.l. The remaining species --- *B. gerstaeckeri*, *B. hortorum*, *B. pascuorum*, *B. pratorum*, *B. psithyrus*, and *B. terrestris/lucorum* experienced linear or quadratic decline in floral resource abundance with increasing elevation. Floral resource abundance was low for all species at elevations above ~1600 m.s.l. Our model explained 67% of total variance (adjusted R^2^ = 0.41). Both lack of management (p << 0.0001) and mowing (p << 0.0001) had a significant positive effect on mean flower cover relative to the reference level of grasing management. The effect of year was also significant, with 2010 (p << 0.0001) and 2011 (p << 0.0001) exhibiting higher mean flower cover than 2012. See Supplemental Material for tabulated model output and diagnostics.

@@

GDM analysis reproduced the finding that overall β-diversity was dominated by floral turnover, but it also revealed a strongly nonlinear response of floral and interaction β-diversity to elevation **(Figure 4A)**. Turnover was high from ~600-1000 m, representing the transition from valley to lower slopes, and then leveled off from ~1000-1500 m. At around 1500 m --- the elevation of the tree line --- turnover accelerated sharply and remained steep for the remainder of the elevation gradient. In contrast, bumble bee turnover was modest and exhibited no marked thresholds. Our GDMs explained 65% of total deviance in interaction β-diversity, 74% of floral β-diversity, and 27% of bumble bee β-diversity. The conditional effect of elevation was significant (p ~ 0) in all models, and its importance was 1-2 orders of magnitude higher that of sampling frequency and management **(Figure 4B)**. Nevertheless, the conditional effect of sampling frequency was significant (p ~ 0) in the interaction β-diversity model, and mowing management was significant (p ~ 0) in the floral β-diversity model.

@@

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective [@Soroye2020-zb], but they also provide a model system in which to study the basic process of community assembly under climate change, a phenomenon of both historical interest and prospective urgency. The central finding of our study, that elevation gradients are characterized by complex and nonlinear responses of species, communities, and biotic interactions --- punctuated by the tree line --- is almost certainly not an idiosyncrasy of bumble bees and their floral hosts but rather a pattern emerging from fundamental physical constraints that should be expected to obtain broadly across the biotic communities of mountains worldwide. Species interactions along tree line ecotones, therefore, deserve special consideration in research and conservation management, since climate-induced range shifts at tree lines should be expected to have disproportionate ecological consequences relative to range shifts at other elevation bands.

<!-- Retreating to higher elevation in response to climatic warming is not a new phenomenon in the evolutionary history of bumble bees [@Hines2008-er; @Martinet2018-rb]. Nevertheless, the rapid pace of recent climate change, together with the interacting effects of human land use, are predicted to [@Soroye2020-zb; @Sirois-Delisle2018-vi]  -->

<!-- - What do these findings tell use about climate change and range shifts in mountain ecosystems more generally? -->
<!-- though as a matter of applied conservation it is perhaps doubtful that local interventions could forestall for long the consequences of a process occurring at a global scale. -->

@@

We thank D.J. McNeil for helpful conversations during data analysis.

@@

**Figure 1:** Study sites in the Berchtesgadener Alps plotted over Google Earth imagery. Sites above the tree line are depecited in red and sites below the tree line in yellow. The elevation of each site is given in meters above sea level.

@@

**Figure 4:** GDM splines of species-level and interaction-level β-diversity in response to elevation (A) and tabulated model output showing variable importance and significance for elevation, sampling frequency, and site management. The maximum height of each spline (A) represents the partial effect of elevation on β-diversity (its effect when covariates are held constant), and the shape of each spline represents to the rate of species turnover as it varies along the elevation gradient, with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Rug plots depict the locations of sites along the elevation axis and are color-coded to indicate whether each site fell above or below the tree line. Variable importance (B) is mapped to a color ramp, and p-values less than 0.05 are depicted in bold typeface.

>>>>

---
title: "The tree line ecotone punctuates the abundance and turnover of plants, pollinators, and their interactions"

@@

Data are provided as private-for-peer review. This study also uses novel code. All data and code required to reproduce our analyses are archived on Dryad: https://datadryad.org/stash/share/M_4Yt9VVXTRm36FSfwwGzUs2OTXP9Fv52jjETdSu6ys

@@

Mountains are unique ecological theaters in which global climate dynamics interact with local climate gradation to shape the distributions of species and biotic interactions [@Telwala2013-sc; @CaraDonna2014-nq; @Miller-Struttmann2014-wz; Peters2019-km; @Rafferty2020-aj]. The rate of recent climate change, however, is forcing species to adapt rapidly within the constraints of gene pools shaped by cooler climates [@Visser2008-ct]. As temperatures rise, species are expected to track suitable conditions to higher elevations, and such effects are already measurable across a broad range of taxa [@Lenoir2015-tj]. Elevational range shifts, however, are not merely abstract translations along the linear lapse of temperature generated by mountain slopes; they are movements *in situ*, involving the advent, extinction, and reorganization of biological interactions [@Tylianakis2008-fi; @Blois2013-dz] as well as the concomitant variation of abiotic conditions other than temperature [@Korner1995-mh; @Hodkinson2005-af]. Thus, when species move along continuous, linear climate gradients, they can experience punctuated and nonlinear ecological transitions. A salient example is the tree line ecotone that marks the elevational transition between forest and grassland. At this relatively discrete band within a linear elevational temperature gradient, plant communities exhibit rapid species turnover [@Descombes2017-id] and declining diversity [@Becker2007-cg] as abiotic conditions become more extreme away from the shelter of tree canopy [@Slatyer1992-ir].

@@

Although per-sample effort was standardized across all years, the total number of samples (i.e. site-dates) varied: 153 in 2010, 212 in 2011, and 539 in 2012. Since the start of field sampling each year was determined by snowmelt, lower sites were sampled as early as April while higher sites could not be sampled until June or July. This resulted in a temporally staggered pattern of sampling in 2010 and 2012, while in 2011 a combination of early snowmelt and a late start to field activities resulted in sampling that was approximately aligned in time across the whole elevation gradient. See Appendix S1 for details of sampling patterns.

@@

Bumble bee abundance was quantified as the total number of recorded floral visits per bumble bee species per site-date. To weight floral abundance based on the foraging preferences of each bumble bee species in our network, we scored each plant genus in our data set as either visited or not visited by each bumble bee species. We then calculated the total floral resource availability for each bumble bee species for each site-date by summing the observed flower cover of *visited* genera. We opted to work at the genus level based on the reasoning that if a bumble bee species visits one member of a given genus, other members of the same genus should also be considered potential floral hosts. This approach was intended to dampen the effects of false non-detection, particularly for rare bumble bee species whose diet breadth would tend to be underestimated simply due to sparsity of observations [@Williams2005-yp]. We treated the cryptic *B. terrestris/lucorum* species pair as a single morphospecies, since these two species cannot be visually distinguished, and we pooled representatives of the parasitic *Psithyrus* subgenus --- *B. barbutellus*, *B. bohemicus*, *B. campestris*, *B. flavidus*, *B. quadricolor*, and *B. sylvestris* --- into the species group *B. psithryus*.  *B. hypnorum*, *B. humilis*, and *B. lapidarius* were extremely rare in our samples, so we omitted them from abundance analyses but retained them for β-diversity analyses.

Eight plant species --- *Rubus idaeus*, *Rosa canina*, *Juniperus communis*, *Larix decidua*, *Salix* sp., *Caltha palustris*, *Pulmonaria officinalis*, and *Rheum barbarum* --- were recorded in visitation data but not recorded during floral surveying. Each accounted for no more than 4 visits in total over the three years of our study, and they were omitted from the analysis of floral abundance.

@@

We analyzed the abundance of bumble bees and floral resources, respectively, using hierarchical generalized additive models (HGAMs) [@Wood2011-pg; @Pedersen2019-kf]. To focus on elevational patterns, we collapsed the temporal variation of abundance within sites into summary values. For bumble bees, we modeled *peak* abundance at each site as a smooth response to `elevation`, with bumble bee `species` as a smoothing factor, `management` (grazing, mowing, or unmanaged) as a fixed effect, and `species:year` as a random intercept term (exploratory analysis indicated that species varied differently across years). For floral resources, we modeled *mean* flower cover, weighted per bumble bee species (i.e. the sum of *visited* genera), as a smooth response to `elevation`, with `management` as a fixed effect, and `year` and bumble bee `species` as random intercept terms. For the bumble bee model, we used the negative binomial distribution family (log link function), and for the floral model we used a Tweedie family distribution (log link function). Both models follow the "type I" model form described in @Pedersen2019-kf.

@@

For our analyses of species and interaction β-diversity, we pooled all samples (including across years) within sites, since the sparse observations of individual sampling events tended to inflate β-diversity and its variance to a degree that impaired model fitting and interpretation. After pooling, we calculated species (separately for bumble bees and plants) and interaction β-diversity between all pairs of sites using the binary Jaccard index and partitioned total interaction β-diversity into its components of species turnover (changes in species composition) and interaction rewiring (interaction turnover despite species constancy) [@Novotny2009-my; @Poisot2012-fk]. Only plant species that were visited at least once by a bumble bee over the course of the study were included in the calculation of floral β-diversity.

@@

As an initial analysis, we visualized the relationship between site-wise elevation difference and each metric of β-diversity, fitting each response with a binomial regression smooth. To investigate the relationship between elevation and β-diversity more deeply, we performed a second analysis using generalized dissimilarity modeling (GDM) [@Ferrier2007-ji] to analyze the elevational variation in the β-diversity of bumble bees, flora, and (unpartitioned) interactions, respectively. In addition to estimating the relationship between differences in a response variable and differences in a predictor variables, GDM captures the slope of this relationship over the range of the predictor variable, revealing potential variation in the amount of change in the response induced by a given change in the predictor [@Ferrier2007-ji]. This enabled us to ask whether the relationship between β-diversity and elevation exhibits thresholds or other nonlinearities. We also included site management in each model by encoding management classes as a series of binary dummy variables, following @Ferrier2007-ji. To control for potential confounding effects of the number of sampling days per site, sampling days was included as a covariate. The importance of the variables in each model were assessed by a permutation test in which variable importance is defined as the percent loss of explained deviance when a given variable is permuted in isolation [@Fitzpatrick-2021aa].

@@

All analyses were conducted in `R` [@RCoreTeam-2021aa]. Data handling and visualization were performed with the `tidyverse` suite [@Wickham2019-aa]. GAM analyses were performed with packages `mgcv` [@Wood_2017] and `mgcViz` [@Fasiolo2018aa]. β-diversity was calculated with the package `vegan` [@Oksanen_2019] and visualized using the packages `ggplot2` [@Wickham2016-aa] and `ggvegan` [@Simpson2019aa]. Calculation and partitioning of interaction β-diversity were performed with the package `bipartite` [@Dormann2008-aa; Dormann2009-aa]. GDM analysis of β-diversity was performed with the package `gdm` [@Fitzpatrick-2021aa]. Data and code to reproduce these analyses are archived on Dryad: https://datadryad.org/stash/share/M_4Yt9VVXTRm36FSfwwGzUs2OTXP9Fv52jjETdSu6ys.

@@

We recorded a total of 12,918 bumble- bee-flower interactions (excluding males) over the three years of our study. The metaweb across all sites and dates consisted of 16 bumble bee species, 163 plant species (110 genera, 37 families), and 736 unique bumble- bee-plant interaction pairs. *B. pascuorum* ("pasc"), *B. pratorum* ("prat"), *B. soroeensis* ("soro"), *B. terrestris/lucorum* ("telu"), and *B. wurflenii* ("wurf") accounted for the bulk of overall bumble bee abundance, while *B. hortorum* ("hort"), *B. jonellus* ("jone"), *B. psithyrus* ("psit"), *B. monticola* ("mont"), *B. mendax* ("mend"), *B. mucidus* ("muci"), *B. pyrenaeus* ("pyre"), and *B. gerstaeckeri* ("gers") were present at moderate to low abundance. *B. hypnorum*, *B. humilis*, and *B. lapidarius* were recorded only sporadically. See Appendix S1 for a summary of species-wise abundances.

Floral surveying yielded a total of 352 plant species (192 genera and 52 families). Of these, 163 species (111 genera and 37 families) were observed to be visited by bumble bees.

@@

HGAM analysis revealed that the response of peak abundance to elevation differed significantly (p ≤ 0.02) from a flat line for all bumble bee species **(Figure 2A)**. *B. mendax*, *B. monticola*, *B. mucidus*, *B. pyrenaeus*, and *B. soroeensis* exhibited a positive response to elevation and peaked in abundance above the tree line, consistent with their known affinity for high elevation habitats [@Rasmont2010-ci]. In contrast, the abundance of *B. hortorum*, *B. pascorum*, *B. psithyrus*, and *B. terrestris/lucorum* declined with elevation. This effect was especially pronounced in *B. pascuorum*, likely because its exceptionally long nesting cycle is incompatible with the short growing season at high elevations [@Neumayer2009-gx]. *B. gerstaeckeri*, *B. jonellus*, and *B. wurflenii* peaked at mid-elevation below the tree line, while *B. pratorum* was unique in exhibiting a sharp peak of abundance *at* the tree line. Our model explained 71% of total variance (adjusted R^2^ = 0.45). Neither mowing (p = 0.08) nor non-management (p = 0.95) had a significant effect on bumble bee abundance relative to the reference level of grasing management. The random effect of species:year was also nonsignificant (p = 0.52). See Appendix S1 for tabulated model output and diagnostics.

Mean floral resource abundance likewise responded significantly (p << 0.0001) to elevation for all bumble bee species **(Figure 2B)**. *B. jonellus*, *B. mendax*, *B. monticola*, *B. mucidus*, *B. soroeensis*, and *B. wurflenii* exhibited bimodal patterns of flora resource abundance, with peaks around 1000 and 1600 m.s.l, the latter peak corresponding closely to the location of the tree line. *B. pyrenaeus* exhibited a unimodal floral resource pattern peaking near 1600 m.s.l. The remaining species --- *B. gerstaeckeri*, *B. hortorum*, *B. pascuorum*, *B. pratorum*, *B. psithyrus*, and *B. terrestris/lucorum* experienced linear or quadratic decline in floral resource abundance with increasing elevation. Floral resource abundance was low for all species at elevations above ~1600 m.s.l. Our model explained 67% of total variance (adjusted R^2^ = 0.41). Both lack of management (p << 0.0001) and mowing (p << 0.0001) had a significant positive effect on mean flower cover relative to the reference level of grasing management. The effect of year was also significant, with 2010 (p << 0.0001) and 2011 (p << 0.0001) exhibiting higher mean flower cover than 2012. See Appendix S1 for tabulated model output and diagnostics.

@@

GDM analysis revealed a strongly nonlinear response of floral and interaction β-diversity to elevation **(Figure 4A)**. Turnover was high from ~600-1000 m, representing the transition from valley to lower slopes, and then leveled off from ~1000-1500 m. At around 1500 m --- the elevation of the tree line --- turnover accelerated sharply and remained steep for the remainder of the elevation gradient. In contrast, bumble bee turnover was modest and exhibited no marked thresholds. Our GDMs explained 65% of total deviance in interaction β-diversity, 74% of floral β-diversity, and 27% of bumble bee β-diversity. The conditional effect of elevation was significant (p ~ 0) in all models, and its importance was 1-2 orders of magnitude higher that of sampling frequency and management **(Figure 4B)**. Nevertheless, the conditional effect of sampling frequency was significant (p ~ 0) in the interaction β-diversity model, and mowing management was significant (p ~ 0) in the floral β-diversity model.

@@

The outcomes of range shifts in mountain bumble bee and floral communities have obvious significance from a conservation perspective [@Soroye2020-zb], but they also provide a model system in which to study the basic process of community assembly under climate change [@Hoiss2012-sn; @Classen2015-jc]. The central finding of our study, that elevation gradients are characterized by complex and nonlinear responses of species, communities, and biotic interactions --- punctuated by the tree line --- is almost certainly not an idiosyncrasy of bumble bees and their floral hosts but rather a pattern emerging from fundamental physical constraints that should be expected to obtain broadly across the biotic communities of mountains worldwide. Species interactions along tree line ecotones, therefore, deserve special consideration in research and conservation management, since climate-induced range shifts at tree lines should be expected to have disproportionate ecological consequences relative to range shifts at other elevation bands.

@@

We thank the National Park Berchtesgaden for their support and the permission to work in the national park, and we thank the owners of the study sites outside the national park for access to their land. The field work of this study was performed within the framework of the joint research center FORKAST as part of the ‘Bavarian Climate Programme 2020'. D.J. McNeil provided helpful discussion of data analyses.

@@

**Figure 1:** Study sites in the Berchtesgadener Alps plotted over Google Earth imagery (https://earth.google.com/). Sites above the tree line are depecited in red and sites below the tree line in yellow. The elevation of each site is given in meters above sea level.

@@

**Figure 4:** GDM splines of species-level and interaction-level β-diversity in response to elevation (A) and tabulated model output showing variable importance and significance for elevation, sampling frequency, and site management. The maximum height of each spline (A) represents the partial effect of elevation on β-diversity (its effect when covariates are held constant), and the shape of each spline represents to the rate of species turnover as it varies along the elevation gradient, with steeper parts of the curve indicating regions of the gradient over which species/interaction turnover is more rapid. Rug plots depict the locations of sites along the elevation axis and are color-coded to indicate whether each site fell above or below the tree line. Variable importance (B) is mapped to a color ramp, and p-values less than 0.05 are depicted in bold typeface. P-values of 0.00 arise from the discreteness of the permutation test and should be interpreted as arbitrarily small.
